# Insert chapter 2 title here

*Describe the work you have done this week and summarize your learning.*

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

```{r}
date()
lrndata <- read.csv("data/learning2014.csv")

```

1. 
The data contain a condensed version of a full questionnaire dataset (details at https://www.mv.helsinki.fi/home/kvehkala/JYTmooc/JYTOPKYS2-meta.txt). Students in an introductory statistics course have answered a questionnaire regarding their approaches and attitudes towards learning and statistics; additionally, their exam points and other background variables have been collected.

Questionnaire variables: 
Respondents have given their answers on a 1-5 scale. 
From the raw data, certain variables have been combined (and scaled back) in order to measure the respondents' use of the following learning approaches: deep (variable name deep), strategic (stra), and surface (surf). The variable 'attitude' combines questions about respondents' attitude towards statistics.

Other variables:
  gender
  age
  points (note: respondents with 0 points have been removed)

```{r}

dim(lrndata)

```

