# Clustering and classification



```{r}
date()

#library(GGally)
#library(dplyr)
library(MASS)
library(tidyr)
library(corrplot)
library(ggplot2)
data("Boston")

```

## 2. 
The data set describes housing value in suburbs of Boston, Massachusetts. The data set has been put together and analyzed by Harrison & Rubinfeld (1978). The raw data has been collected in ~1970 (Harrison & Rubinfeld, 1978).

Listing of its variables are available [here](https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/Boston.html), for instance. Definitions of how the variables have been collected and calculated can be found in Harrison, D. and Rubinfeld, D.L. (1978) Hedonic prices and the demand for clean air. J. Environ. Economics and Management 5, 81–102. 

```{r}

str(Boston)

```
There are 506 observations and 14 variables. Each observation represents one neighborhood/town. 

In summary, the variables include median values of homes, variables describing housing, infrastructural, environmental and industrial structure, and demographic information on the neighborhood (Harrison & Rubinfeld, 1978). 

```{r}

summary(Boston)

```

The median value of housing seems to be capped at $50.000, which looks like an artificial decision. 

Let's visualize the data set. 

## 3. 

```{r}
# Create scatterplot matrix
pairs(Boston)

```
### Commenting on median value plotted against other factors: 
Not surprisingly, higher-value areas have lower crime rates. The density of residential areas (var zn) and whether tract bounds river (chas [this is not surprising as this is a dummy variable anyway]) do not have a clear effect on housing value. Prices seem to be lower where industrial density is very high, but low industrial density has both high and low-value areas. The situation looks fairly similar with highly polluted areas (one might ask how independent nox and indus are). As average number of rooms in units (rm) grows, so does value -- again, not surprisingly. 
Newest areas are not cheap, but the priciest areas are not new. Both the cheapest and the most expensive area seem to have a short distance to Boston employment centers. Rad -- index of accessibility to radial highways -- seems to have a very binary distribution despite its values varying from 1 to 24. I am not sure how to interpret how tax and value are related (the highest taxes seem to be in cheapest areas, which seems odd from someone coming from a progressive tax state). Pupil-teacher ratio is generally lower in more valuable neighborhoods, as one would expect. Towns with a lower proportion of African-Americans are not among the most expensive areas. There is a high correlation between the price of housing and lower percentage of "lower-status" population (i.e., male workers classified as laborers and adults with no high school education).


```{r}
# Look at correlations
corbos = cor(Boston)
corrplot(corbos)
```
The strongest correlation seems to be between radial highway accessibility and property tax rate. Looking at the scatterplot matrix, there is a highly taxed highly accessible outlier, which may confound the results. 
Distance and pollution have a strong negative correlation: far-away areas have lower pollution.
Nox and indus are also strongly correlated.

## 4. 
```{r}

# Standardize the dataset and print out summaries of the scaled data. How did the variables change? Create a categorical variable of the crime rate in the Boston dataset (from the scaled crime rate). Use the quantiles as the break points in the categorical variable. Drop the old crime rate variable from the dataset. Divide the dataset to train and test sets, so that 80% of the data belongs to the train set. (0-2 points)

boston_scaled <- scale(Boston)
summary(boston_scaled)
```
Columns are scaled by first subtracting column values by the corresponding column mean, and then dividing the difference by standard deviation. This way, the mean for all variables is at 0, and we can compare variables in a more meaningful way.

```{r}

boston_scaled<-as.data.frame(boston_scaled)
boston_scaled$crim <- as.numeric(boston_scaled$crim)
# create a quantile vector of crim and print it
bins <- quantile(boston_scaled$crim)
bins

# create a categorical variable 'crime'
crime <- cut(boston_scaled$crim, breaks = bins, include.lowest = TRUE, label=c("low","med_low","med_high","high"))

# look at the table of the new factor crime
table(crime)

# remove original crim from the dataset
boston_scaled <- dplyr::select(boston_scaled, -crim)

# add the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)

# number of rows in the Boston dataset 
n <- nrow(boston_scaled)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)

# create train set
train <- boston_scaled[ind,]

# create test set 
test <- boston_scaled[-ind,]

```

## 5. 

```{r}
# Fit the linear discriminant analysis on the train set. Use the categorical crime rate as the target variable and all the other variables in the dataset as predictor variables. Draw the LDA (bi)plot. (0-3 points)

# linear discriminant analysis
lda.fit <- lda(crime ~ ., data = train)

# print the lda.fit object
lda.fit

# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(train$crime)

# plot the lda results
plot(lda.fit, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 2.5)


```


## 6.

```{r}
# Save the crime categories from the test set and then remove the categorical crime variable from the test dataset. Then predict the classes with the LDA model on the test data. Cross tabulate the results with the crime categories from the test set. Comment on the results. (0-3 points)

# save the correct classes from test data
correct_classes <- test$crime

# remove the crime variable from test data
test <- dplyr::select(test, -crime)

# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)
```


```{r}
# cross tabulate the results
results <- table(correct = correct_classes, predicted = lda.pred$class)

# calculate model precision & sensitivity
table(correct = correct_classes, predicted = lda.pred$class)
1-diag(results)/colSums(results) # Precision: fraction of false positives
1-diag(results)/rowSums(results) # Sensitivity: fraction of false negatives

```
The model is best at predicting high crime areas. No areas were misclassified as high, although one high area was falsely classified as med_high. The model had more difficulties with classes med_high and med_low; many med_high classes were incorrectly classified as med_low, making 45% of med_low classifications incorrect and greatly reducing the sensitivity of med_high classification. Areas with a low crime rate were better classified, with about 80% precision and 26% sensitivity.

## 7.
```{r}
# Make another scaled Boston data set
boston_scaled <- scale(Boston)

# Distances between observations
summary(dist(boston_scaled))

```

An ideal grouping has a minimal within-group sum of squares ("WGSS") over all variables. Let's plot these for one to ten-group solutions to see which solution has the smallest WGSS.

```{r}
set.seed(123)

# determine the number of clusters
k_max <- 10

# calculate the total within sum of squares
twcss <- sapply(1:k_max, function(k){kmeans(Boston, k)$tot.withinss})

# visualize the results
df<- data.frame(wgss=c(twcss),groups=1:k_max)
ggplot(data = df, aes(x=groups, y=wgss)) + geom_line() + scale_x_continuous(breaks = 1:10)

```
There are two "elbows" in the plot, at 3 and 8 groups. Let's investigate those further. 

```{r}
set.seed(123)

# k-means clustering
km3 <- kmeans(boston_scaled, centers = 3)

# plot the Boston dataset with clusters
pairs(boston_scaled, col = km3$cluster)

# k-means clustering
km8 <- kmeans(boston_scaled, centers = 8)

# plot the Boston dataset with clusters
pairs(boston_scaled, col = km8$cluster)

```
It's difficult to point out any single variable based on which the groups have been clustered (which is a good sign, I suppose).  It's likely that km3 makes a better model of the data as groups 4-7 actually make the model worse; km8 might just overfit the data by dividing groups into unnecessary subgroups. 


